Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-05 resubmission/01-introduction.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

The justification of economic inequality is a topic with a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986]. Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" [-@lane_market_1986, p.384], proposing the distinction between market and political justice. Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on individual merit. Part of this agenda has focused on analyzing the extent to which individuals consider as just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015], usually under the label of market justice preferences. The empirical evidence to date suggests that a significant portion of the population justifies inequality in these areas [@koos_moral_2019], and that such preferences vary according to objective socioeconomic factors and countries contextual characteristics [@lindh_public_2015; @busemeyer_skills_2014]. Within this framework, we argue that there are two relevant factors that are usually sidelined in this literature and that are key to understand market justice preferences: the perception of economic inequality and the perception of meritocracy.

>>>>

The justification of economic inequality is a topic with a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986]. Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" [-@lane_market_1986, p.384], proposing the distinction between market and political justice. Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on individual merit. Part of this agenda has focused on analyzing the extent to which individuals consider as just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015], usually under the label of market justice preferences. Although these services can be considered fundamental rights of citizenship, existing evidence indicates that a substantial share of the population legitimizes inequality in these domains, and that such preferences are shaped by individuals' socioeconomic characteristics as well as the broader contextual features of their countries [@koos_moral_2019; @lindh_public_2015; @busemeyer_skills_2014]. Within this framework, we argue that there are two relevant factors that are usually sidelined in this literature and that are key to understand market justice preferences: the perception of economic inequality and the perception of meritocracy.

### 2025-08-05 resubmission/02-antecedents.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice. The author defines market justice as a system of "earned deserts", whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility [@kluegel_legitimation_1999; @lindh_public_2015].

@@

Assessing the perception of inequality in empirical research is an ongoing challenge. Specifically in survey research, the assessment of perceived inequality through Likert-type items such as "income differences are too large" have been questioned as they captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019]. As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies. To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and perceived estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017]. Through asking perceived salaries for occupations in the extremes of the status continuum (such as a CEO and a manual worker), an indirect measure of perceived inequality is built by the ratio of the high and the low status occupation. This type of measure provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015]. Evidence to dat shows that perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.

@@

Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023]. Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020]. This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives. Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@lopez-roldan_comparative_2021; @torche_intergenerational_2014a]. While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@espinoza_movilidad_2014].

>>>>

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice. The author defines market justice as a system of "earned deserts", whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement and procedures [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility [@kluegel_legitimation_1999; @lindh_public_2015].

@@

Assessing the perception of inequality in empirical research is an ongoing challenge. Specifically in survey research, the assessment of perceived inequality through Likert-type items such as "income differences are too large" have been questioned as they captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019]. As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies. To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and perceived estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017]. Through asking perceived salaries for occupations in the extremes of the status continuum (such as a CEO and a manual worker), an indirect measure of perceived inequality is built by the ratio of the high and the low status occupation. This type of measure provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015]. Evidence to date shows that perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.

@@

Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023]. Following the neoliberal reforms of the 1980s introduced during the civil-military dictatorship (1973-1989), the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020]. This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives. Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@lopez-roldan_comparative_2021; @torche_intergenerational_2014a]. While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@espinoza_movilidad_2014].

### 2025-08-05 resubmission/03-methods.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic. This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest. Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals. Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response. Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.

>>>>

The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic. This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest. Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals. Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response. Further details on sampling, attrition, and weighting procedures are available at [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/), and the dataset is publicly accessible at [https://dataverse.harvard.edu/dataverse/elsoc](https://dataverse.harvard.edu/dataverse/elsoc).

### 2025-08-05 resubmission/05-discussion.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

The first hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021]. Interestingly, intraindividual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects. While an increase in the perception that rewards are based on individual effort is associated with stronger preferences for market fairness, an increase in the belief that rewards come from talent is linked to lower support for such principles. One possible explanation is that effort is generally perceived as a controllable and malleable trait, which aligns with the normative foundations of market fairness. Conversely, talent may be viewed as an innate and less alterable attribute, making talent-based reward allocation seem less fair or legitimate. Moreover, greater exposure to real-life situations where outcomes are strongly determined by inherent characteristics, rather than effort or hard work, may lead people to question the fairness of market-based reward systems.

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to strengthen when perceptions of economic inequality were particularly high. While this moderation is not significant for effort-related effects, it is significant for talent in the within- and between-individual effects. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social movements—particularly those that emerged in 2019—on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice. The main effect of time suggests that, on average, support for market justice increased after 2019. However, the results do not reveal any significant interactions between distributive mechanisms and market justice attitudes, indicating that the 2019 social mobilizations did not substantially alter the underlying associations between these variables. This may suggest a certain stability or resilience in the normative frameworks that shape individuals’ evaluations of justice, even in the context of major collective political events. Alternatively, it is possible that while the protests challenged institutional legitimacy, they did not fundamentally shift individuals’ views regarding how rewards should be allocated in society.

>>>>

The first set of hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].  Such a relationship reflects a moral economy in which perceptions of systemic unfairness undermine the legitimacy of existing distributions and fuel demands for greater equity. In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines—potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system. However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant. Therefore, we cannot conclude that individuals who, on average, perceive higher inequality are systematically less supportive of market justice. As a result, hypothesis $H1_{a}$ does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that meritocratic perceptions—particularly those emphasizing individual effort—were associated with stronger support for market-based distribution systems. Individuals who believed that success is primarily achieved through personal effort were more likely to justify unequal access to core social services based on income, interpreting such disparities as outcomes of individual merit rather than systemic injustice ($H2_{a}$). This finding aligns with previous research showing that meritocratic narratives serve as moral justifications that legitimize social stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions operate symbolically to reinforce structural inequalities by reducing support for redistributive policies, framing inequality as both fair and deserved. Consistent with prior work by @castillo_socialization_2024 on Chilean students, the findings suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure. This mechanism is particularly salient in neoliberal contexts like Chile, where market logics heavily shape social attitudes [@canalesceron_sujeto_2021].

Interestingly, intra-individual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects. While increases in the belief that rewards are based on individual effort are associated with stronger preferences for market justice, increases in the belief that rewards derive from talent are linked to weaker support for such principles. One possible explanation is that effort is generally viewed as a controllable and malleable trait, aligning with the normative foundations of market fairness. In contrast, talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate. Furthermore, increased exposure to real-world scenarios in which outcomes are clearly shaped by inherent traits rather than hard work may lead individuals to question the fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions of inequality in the relationship between meritocratic perceptions and preferences for market justice. The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high—contradicting our initial expectations ($H3_{a}$ and $H3_{b}$). Although this moderating effect is not statistically significant for effort-related meritocratic perceptions, it is significant for talent-related perceptions at both the within- and between-individual levels. This finding runs counter to our hypotheses, as it suggests that the effect of meritocratic perceptions—specifically, talent-based meritocracy—intensifies as perceptions of inequality increase. However, this interpretation requires caution. The main effect of perceived inequality on support for market justice is negative. Thus, as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger. This suggests a more nuanced dynamic: meritocratic perceptions, particularly those emphasizing talent, may serve as a stronger justificatory mechanism for market-based inequalities among individuals who are less inclined to perceive inequality as a problem. In other words, among those who do not strongly perceive systemic disparities, meritocratic narratives may play a more influential role in legitimizing unequal outcomes.

Finally, the fourth hypothesis examined the potential impact of recent social mobilizations—particularly those that erupted in 2019—on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice. The main effect of time indicates that, on average, support for market justice increased following the 2019 protests. However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that the social mobilizations did not substantially modify the underlying associations between these variables. As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the data. This may point to a certain stability—or even resilience—in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events. Alternatively, while the protests may have eroded trust in institutional arrangements or governance, they may not have fundamentally altered individuals’ beliefs about how rewards and resources should be allocated in society.

### 2025-08-05 resubmission/blinded-paper-resubmit.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
# author:
#   - name: Juan Carlos Castillo
#     corresponding: true
#     orcid: 0000-0003-1265-7854
#     email: juancastillov@uchile.cl
#     affiliations:
#       - ref: 1
#       - ref: 2
#       - ref: 3
#   - name: Andreas Laffert
#     orcid: 0000-0002-9008-2454
#     email: alaffertt@estudiante.uc.cl 
#     affiliations:
#       - ref: 4
#   - name: Kevin Carrasco
#     orcid: 0000-0002-9348-8361
#     email: kevin.carrasco@ug.uchile.cl
#     affiliations:
#       - ref: 2
#   - name: Julio Iturra-Sanhueza
#     orcid: 0000-0003-1993-3479
#     email: julioiturrasanhueza@gmail.com 
#     affiliations:
#       - ref: 5     
# affiliations: 
#   - id: 1
#     name: Departamento de Sociología, Universidad de Chile
#   - id: 2
#     name: Centro de estudios del conflicto y cohesión social (COES)
#   - id: 3
#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
#   - id: 4
#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
#   - id: 5
#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
 Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services. In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services. From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences. This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. The findings show that greater perceived inequality is associated with lower market justice preferences. However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities. In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness. These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks.
  \newline
  **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
# author:
#   - name: Juan Carlos Castillo
#     corresponding: true
#     orcid: 0000-0003-1265-7854
#     email: juancastillov@uchile.cl
#     affiliations:
#       - ref: 1
#       - ref: 2
#       - ref: 3
#   - name: Andreas Laffert
#     orcid: 0000-0002-9008-2454
#     email: alaffertt@estudiante.uc.cl 
#     affiliations:
#       - ref: 4
#   - name: Kevin Carrasco
#     orcid: 0000-0002-9348-8361
#     email: kevin.carrasco@ug.uchile.cl
#     affiliations:
#       - ref: 2
#   - name: Julio Iturra-Sanhueza
#     orcid: 0000-0003-1993-3479
#     email: julioiturrasanhueza@gmail.com 
#     affiliations:
#       - ref: 5     
# affiliations: 
#   - id: 1
#     name: Departamento de Sociología, Universidad de Chile
#   - id: 2
#     name: Centro de estudios del conflicto y cohesión social (COES)
#   - id: 3
#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
#   - id: 4
#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
#   - id: 5
#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
 Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services. In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services. From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences. This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. The findings show that greater perceived inequality is associated with lower market justice preferences. However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities. In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness. These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
  \newline
 **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
